Supporting Information for

### 2021-03-07 Manuscript.Rmd
fill in supplement

---
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: no
    number_sections: false
    keep_tex: true
    includes:
      in_header: "preamble.tex"
    citation_package: natbib
    extra_dependencies: ["float"]
bibliography: "references.bib"
biblio-style: "genetics"
title: Linkage disequilibrium in conserved domains reveals opposing selective interactions in genes
author:
header-includes:
- |
  ```{=latex}
  \usepackage{authblk}
  \author{Aaron P. Ragsdale \\ aaronpeaceragsdale@gmail.com}
  \affil{Department of Human Genetics, McGill University, Montreal, Canada \\
         National Laboratory of Genomics for Biodiversity, Irapuato, Mexico \\
         }
  ```
date: \today
---

@@

Most new mutations that affect fitness are deleterious and tend to be
eliminated from a population. The amount of time that a deleterious mutation
segregates depends on the strength of selection against genomes that carry it,
with very damaging mutations kept at low frequencies and purged relatively
rapidly. In the time between mutation and fixation or loss, selected variants,
both beneficial and damaging, can dramatically impact patterns of variation in
nearby linked regions. This distortion away from neutral expectations has been
widely documented in sequencing data, but questions remain about the primary
mode of interactions between multiple linked variants and their joint effects
on genome-wide patterns of diversity.

>>>>

---
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: no
    number_sections: false
    keep_tex: true
    includes:
      in_header: "preamble.tex"
    citation_package: natbib
    extra_dependencies: ["float"]
bibliography: "references.bib"
biblio-style: "genetics"
title: Can we distinguish modes of selective interactions using linkage disequilibrium?
author:
header-includes:
- |
  ```{=latex}
  \usepackage{authblk}
  \author{Aaron P. Ragsdale \\ aaronpeaceragsdale@gmail.com}
  \affil{Department of Human Genetics, McGill University, Montreal, Canada \\
         National Laboratory of Genomics for Biodiversity, Irapuato, Mexico \\
         }
  ```
date: \today
---

@@

Most new mutations that affect fitness are deleterious and tend to be
eliminated from a population. The amount of time that a deleterious mutation
segregates depends on the strength of selection against genomes that carry it,
with very damaging mutations kept at low frequencies and purged relatively
rapidly. In the time between mutation and fixation or loss, selected variants,
both beneficial and damaging, can dramatically impact patterns of variation in
nearby linked regions. This distortion away from neutral expectations is now
empirically documented using sequencing data from an ever-growing set of study
systems, but questions remain about the primary mode of interactions between
multiple linked variants and their joint effects on genome-wide patterns of
diversity.

### 2021-03-07 SupportingInformation.Rmd
fill in supplement

# Supporting Information

@@

## The diffusion equation and moment system for the two-locus sampling distribution

The two-locus diffusion equation with additive selection was first described by
@Kimura1955-qe studied extensively in the 1960s and 70s, including by
@Hill1966-gv and @Ohta1969-ie. The continuous distribution $\psi(x_1, x_2,
x_3)$ of haplotype frequencies in a population, where $x_1$ is the frequency of
$AB$, $x_2$ of $Ab$, and $x_3$ of $aB$, is governed by the multi-dimensional
Fokker-Planck equation:
\begin{align} \label{eq:diffeq}
\frac{\partial \psi}{\partial \tau} = &
\frac{1}{2}\mathop{\sum\sum}_{1\leq i, j \leq 3} 
\frac{\partial^2}{\partial x_i \partial x_j} 
\left[\frac{x_i(\delta_{i=j}-x_j)\psi}{\nu(\tau)}\right] \\\nonumber
& -\frac{\rho}{2}\left(-\frac{\partial}{\partial x_1} D\psi
  + \frac{\partial}{\partial x_2} D\psi
  + \frac{\partial}{\partial x_3} D\psi\right) \\\nonumber
& - \frac{\gamma_A}{2}\left[
  \frac{\partial}{\partial x_1} x_1(1-x_1-x_2)\psi
  + \frac{\partial}{\partial x_2} x_2(1-x_1-x_2)\psi
  - \frac{\partial}{\partial x_3} x_3(x_1+x_2)\psi
  \right] \\\nonumber
& -\frac{\gamma_B}{2}\left[
  \frac{\partial}{\partial x_1} x_1(1-x_1-x_3)\psi
  - \frac{\partial}{\partial x_2} x_2(x_1+x_3)\psi
  + \frac{\partial}{\partial x_3} x_3(1-x_1-x_3)\psi
  \right].
\end{align}
$D$ is the standard covariance measure of linkage disequilibrium, \[D=x_1 -
(x_1+x_2)(x_1+x_3) = x_1 x_4 - x_2 x_3,\] $\gamma_A$ and $\gamma_B$ are the
scaled selection coefficients at the left and right locus, and $\rho$ is the
scaled recombination rate between the two loci. Time $\tau$ is measured in
$2N_e$ generations.

@@

- instead of (numerically) solving for $\psi$ and using \@ref(eq:multinomial},
  we can apply the multinomial sampling formula directly to the PDE
- what we get instead is a system of differential equations on the entries of
  $\Psi_n$
- we can then solve for these noncanonical moments of $\psi$, which are the
  objects of interest in any case, and avoid having to solve this numerical PDE
  with unpleasant boundary conditions
- the same system can be found by deriving recursion from considering tracking
  $n$ lineages within the population, in the vein of @Wright1931-xx

### Drift, mutation, recombination, and moment closure

- point to supplement of @Ragsdale2019-nt

### Selection

## Data analysis

### DFE for missense and LOF variants

@@

### Multinucleotide mutations and positive LD between linked synonymous variants

@@

In fitting the LD decay of $\sigma_d^1$, the best fit parameters were
$A=0.132$, and $\lambda=0.0103$. An exponential scaling of $0.01$ implies that
the vast majority of new mutations do not occur via MNMs for distances greater
than 200 bp, though a substantial fraction ($10-15\%$) occur via MNMs for very
tightly linked loci with distances on the order $0-50$ base pairs. It is
important to note that this does not mean that $10-15\%$ of new mutations occur
via MNMs, since this fraction is contitioned on two mutations occurring at
short distances.

### Grouping Thousand Genomes populations based on clustering

Testing grouping populations that cluster together as a single population.

- Does not reduce CIs

Citation for Alex's work:
\url{https://www.abstractsonline.com/pp8/#!/9070/presentation/2384}

>>>>

@@

# The diffusion equation and moment system for the two-locus sampling distribution

The two-locus diffusion equation with additive selection was first described by
@Kimura1955-qe studied extensively in the 1960s and 70s, including by
@Hill1966-gv and @Ohta1969-ie. The continuous distribution $\psi(x_1, x_2,
x_3)$ of haplotype frequencies in a population, where $x_1$ is the frequency of
$AB$, $x_2$ of $Ab$, and $x_3$ of $aB$, is governed by the multi-dimensional
Fokker-Planck equation:
\begin{align} \label{eq:diffeq}
\frac{\partial \psi}{\partial \tau} = &
\frac{1}{2}\mathop{\sum\sum}_{1\leq i, j \leq 3} 
\frac{\partial^2}{\partial x_i \partial x_j} 
\left[\frac{x_i(\delta_{i=j}-x_j)\psi}{\nu(\tau)}\right] \\\nonumber
& -\frac{\rho}{2}\left(-\frac{\partial}{\partial x_1} D\psi
  + \frac{\partial}{\partial x_2} D\psi
  + \frac{\partial}{\partial x_3} D\psi\right) \\\nonumber
& - \frac{\gamma_A}{2}\left[
  \frac{\partial}{\partial x_1} x_1(1-x_1-x_2)\psi
  + \frac{\partial}{\partial x_2} x_2(1-x_1-x_2)\psi
  - \frac{\partial}{\partial x_3} x_3(x_1+x_2)\psi
  \right] \\\nonumber
& -\frac{\gamma_B}{2}\left[
  \frac{\partial}{\partial x_1} x_1(1-x_1-x_3)\psi
  - \frac{\partial}{\partial x_2} x_2(x_1+x_3)\psi
  + \frac{\partial}{\partial x_3} x_3(1-x_1-x_3)\psi
  \right].
\end{align}
$D$ is the standard covariance measure of linkage disequilibrium, \[D=x_1 -
(x_1+x_2)(x_1+x_3) = x_1 x_4 - x_2 x_3,\] $\gamma_A$ and $\gamma_B$ are the
scaled selection coefficients at the left and right locus, and $\rho$ is the
scaled recombination rate between the two loci. Time $\tau$ is measured in
$2N_e$ generations, and $\nu(\tau)$ is the population size relative
to the ancestral size or some reference size at time $\tau$.

@@

In the method-of-moments approach, instead of solving the differential equation
for $\psi$, we instead integrate both sides of the differential equation
against the multinomial sampling function for a given sampling configuration
$(i, j, k)$. On the left side, we get $\partial_t \Psi_n(i, j, k)$, and on the
right we obtain, after some simple integration by parts and somewhat tedious
simplification, terms for drift, recombination, and selection that can be
written as sparse linear operators of $\Psi_n$. Written compactly, this takes
the form
\begin{equation}
\partial_\tau \Psi_n = 
\frac{1}{2\nu(\tau)}\mathcal{D}_{n}\Psi_n
+ \frac{\rho}{2}\mathcal{R}_{n}\Psi_n
+ \frac{\theta}{2}\mathcal{U}_{n}\Psi_n
+ \mathcal{S}_{n, \mathbf{\gamma}, \mathbf{h}}\Psi_n.
\end{equation}.

Alternatively, we arrive at this same linear system of equations by considering
tracking the expected sampling distribution over $n$ lineages within the full
population and how that changes over time by drawing lineages from one
generation to the next in the style of @Wright1931-wy. Both @Jouganous2017-pq,
for the single-locus SFS, and @Ragsdale2019-nt drew this connection in detail,
so I refer readers to that previous work for a fuller description of those
derivations and discussion. In the next section I repeat the results for
$\mathcal{D}$, $\mathcal{R}$, and $\mathcal{U}$, briefly describe the moment
closure approximation (which is the same as presented in @Ragsdale2019-nt), and
then describe the selection operator $\mathcal{S}$ for selection with
epistasis, both with and without dominance.

## Drift, mutation, recombination, and moment closure

### Drift

Drift for an entry $(i, j, k)$ depends only on $\Psi_n$ and therefore closes.
The entries of $\mathcal{D}$ are found by considering the possibility of a
coalescence event occurring within a given generation within $n$ lineages in
the full population. If $n \ll N$, we can safely assume that at most a single
such event occurs in any given generation.

\begin{align}
\mathcal{D}_{n}(i, j, k)\Psi_n =
& (i - 1) (n - i - j - k + 1) \Psi_n(i - 1, j, k) \\\nonumber
& + (i + 1) (n - i - j - k - 1) \Psi_n(i + 1, j, k) \\\nonumber
& + (i - 1) (k + 1) \Psi_n(i - 1, j, k + 1) \\\nonumber
& + (i + 1) (k - 1) \Psi_n(i + 1, j, k - 1) \\\nonumber
& + (i - 1) (j + 1) \Psi_n(i - 1, j + 1, k) \\\nonumber
& + (i + 1) (j - 1) \Psi_n(i + 1, j - 1, k) \\\nonumber
& + (j - 1) (n - i - j - k + 1) \Psi_n(i, j - 1, k) \\\nonumber
& + (j + 1) (n - i - j - k - 1) \Psi_n(i, j + 1, k) \\\nonumber
& + (j - 1) (k + 1) \Psi_n(i, j - 1, k + 1) \\\nonumber
& + (j + 1) (k - 1) \Psi_n(i, j + 1, k - 1) \\\nonumber
& + (k - 1) (n - i - j - k + 1) \Psi_n(i, j, k - 1) \\\nonumber
& + (k + 1) (n - i - j - k - 1) \Psi_n(i, j, k + 1) \\\nonumber
& - 2\left( i ( n - i - j - k) + i k + i j + j (n - i - j - k) + j k + k(n - i - j - k)\right)\Psi_n(i, j, k)
\end{align}

### Recombination

If a lineage in our sample of size $n$ recombines in a given generation, which
occurs with probability $nr$, we need to draw an extra lineage from the full
population for it to recombine with. This means we need $\Psi_{n+1}$ in the
previous generation. After drawing that extra lineage, $\Psi_n$ changes as we
draw one of the two recombinant types (each with probability $1/2$$ instead of
the lineage that was chosen to recombine.

\begin{align}
\mathcal{R}_{n}(i, j, k)\Psi_n = &
\frac{(i + 1)(n - i - j - k + 1)}{n + 1}\Psi_{n+1}(i + 1, j - 1, k) \\\nonumber
& + \frac{(i + 1)(n - i - j - k + 1)}{n + 1}\Psi_{n+1}(i + 1, j, k - 1) \\\nonumber
& + \frac{(j + 1)(k + 1}{n + 1)}\Psi_{n+1}(i - 1, j + 1, k + 1) \\\nonumber
& + \frac{(j + 1)(k + 1}{n + 1)}\Psi_{n+1}(i, j + 1, k + 1) \\\nonumber
& - \frac{(i + 1)(n - i - j - k)}{n + 1}\Psi_{n+1}(i + 1, j, k) \\\nonumber
& - \frac{(j + 1) k}{n + 1}\Psi_{n+1}(i, j + 1, k) \\\nonumber
& - \frac{j (k + 1)}{n + 1}\Psi_{n+1}(i, j, k + 1) \\\nonumber
& - \frac{i (n - i - j - k + 1}{n + 1}\Psi_{n+1}(i, j, k)
\end{align}

### Mutation

We assume an infinite sites mutation (ISM) model where new mutations occur at
previously unmutated loci. In the two-locus ISM model, two-locus pairs of
variable loci arise when a mutation occurs at one locus when the other locus is
already variable. Thus, new mutations at the $B/b$ locus occur against the
single-locus allele frequency distribution $\Phi_{n,A}$, and new mutations at
the $A/a$ locus occur against $\Phi_{n, B}$, which are found via the single-locus
system from @Jouganous2017-pq.

\begin{align}
\mathcal{U}_{n}(i, j, k)\Psi_n = & 
(j + 1)\frac{\theta_B}{2} \Phi_{n, A}(j+1) \delta_{i=1, k=0} \\\nonumber
& + (n - j)\frac{\theta_B}{2} \Phi_{n, A}(j) \delta_{i=0, k=1} \\\nonumber
& + (i + 1)\frac{\theta_A}{2} \Phi_{n, B}(i+1) \delta_{i=1, j=0} \\\nonumber
& + (n - i)\frac{\theta_A}{2} \Phi_{n, B}(i) \delta_{i=0, j=1} \\\nonumber
\end{align}

### Jackknife moment closure approximation

We use a jackknife approximation to write the entries of $\Psi_{n+1}$ and
$\Psi_{n+2}$ as linear combinations of entries in $\Psi_n$. The general
strategy is to assume the underlying continuous distribution $\psi(x, y, z)$
can be approximated locally as a quadratic, and then use entries in $\Psi_n$
that are close in frequency to a given entry in $\Psi_{n+l}$ to estimate the
coefficients of that quadratic using the multinomial sampling formula. Then
this quadratic local approximation to $\psi$ can be used to compute
$\Psi_{n+l}(i, j, k)$ using Eq. \@ref(eq:multinomial). Readers should refer to
section S1.3.5 in the Supporting material for @Ragsdale2019-nt for details.

## Selection

First consider the case of no dominance, so that the haplotypes $Ab$, $aB$,
and $AB$ have selection coefficients $s_{Ab}$, $s_{aB}$, and $s_{AB}$,
respectively. Note that the case with $s_{AB} = s_{Ab} + s_{aB}$ implies no
epistasis between the $A/a$ and $B/b$ loci. Here, we assume all selection
coefficients are negative. In a given generation, a selection event could
occur in which a haplotype is rejected (selected against) with probability
proportional to its selection coefficient, $-s$. We then draw an extra lineage
from the full population to replace that rejected lineage.

For example, the probability that an $AB$ haplotype is selected against and
replaced by an $Ab$ haplotype is \[-ns_{AB} \frac{i}{n+1}{j+1}{n}\Psi_{n+1}(i,
j + 1, k),\] where the additional $j + 1$ lineage in a sample of size $n + 1$
accounts drawing that extra $Ab$ haplotype. Taking all such selective events
together, for additive selection we get
\begin{align}
\mathcal{S}_{n}(i, j, k)\Psi_n = &
\frac{i+1}{n+1}\left(-s_{AB}(n-i) + s_{Ab}j
+ s_{aB}k\right)\Psi_{n+1}(i+1, j, k) \\\nonumber
& +\frac{j+1}{n+1}\left(s_{AB}i - s_{Ab}(n-j)
+ s_{aB}k\right)\Psi_{n+1}(i, j+1, k) \\\nonumber
& +\frac{k+1}{n+1}\left(s_{AB}i + s_{Ab}j
- s_{aB}(n-k)\right)\Psi_{n+1}(i, j, k+1) \\\nonumber
& + \frac{n-i-j-k+1}{n+1}\left(s_{AB}i + s_{Ab}j
+ s_{aB}k\right)\Psi_{n+1}(i, j, k)
\end{align}

For a general diploid selection model, the idea is nearly the same, but we need
to draw an extra lineage to determine the fitness of a diploid individual. For
example, the probability that an $AB$ haplotype is paired with an additional
lineage $Ab$ and selected against, and then replaced by an $aB$ haplotype is
\[-ns_{AB/Ab}\frac{i}{n+2}\frac{j+1}{n+1}\frac{k+1}{n}\Psi_{n+2}(i, j+1,
k+1).\] There are now many more possible selective events to consider, but
after accounting for all possible diploid pairs and replacements (90 in total)
and simplifying, we find
\begin{align}
\mathcal{S}_{n}(i, j, k)\Psi_n = &
\frac{n-i-j-k+2}{n+2}\frac{n-i-j-k+1}{n+1}\left(
  s_{AB/ab}i 
  + s_{Ab/ab}j 
  + s_{aB/ab}k
\right)\Psi_{n+2}(i,j,k) \\\nonumber
& + \frac{i+1}{n+2}\frac{n-i-j-k+1}{n+1}(
  s_{AB/AB}i
  + s_{AB/Ab}j
  + s_{AB/aB}k
  + s_{Ab/ab}j \\\nonumber & \hspace{20pt}
  + s_{aB/ab}k
  - s_{AB/ab}(n + j + k)
)\Psi_{n+2}(i+1, j, k) \\\nonumber
& + \frac{i+2}{n+2}\frac{i+1}{n+1}(
  s_{AB/Ab}j
  + s_{AB/aB}k
  + s_{AB/ab}(n-i-j-k)
  - s_{AB/AB}(n-i)
)\Psi_{n+2}(i+2, j, k) \\\nonumber
& + \frac{i+1}{n+2}\frac{j+1}{n+2}(
  s_{AB/AB}i
  + s_{AB/aB}k
  + s_{AB/ab}(n-i-j-k)
  + s_{Ab/Ab}j \\\nonumber & \hspace{20pt}
  + s_{Ab/aB}k
  + s_{Ab/ab}(n-i-j-k)
  - s_{AB/Ab}(2n-i-j)
)\Psi_{n+2}(i+1, j+1, k) \\\nonumber
& + \frac{i+1}{n+2}\frac{k+1}{n+1}(
  s_{AB/AB}i
  + s_{AB/Ab}j
  + s_{AB/ab}(n-i-j-k)
  + s_{Ab/aB}j \\\nonumber & \hspace{20pt}
  + s_{aB/aB}k
  + s_{aB/ab}(n-i-j-k)
  - s_{AB/aB}(2n-i-k)
)\Psi_{n+2}(i+1, j, k+1) \\\nonumber
& + \frac{j+1}{n+2}\frac{n-i-j-k+1}{n+1}(
  s_{AB/Ab}i
  + s_{AB/ab}i
  + s_{Ab/Ab}j
  + s_{Ab/aB}k \\\nonumber & \hspace{20pt}
  + s_{aB/ab}k
  -s_{Ab/ab}(n+i+k)
)\Psi_{n+2}(i, j+1, k) \\\nonumber
& + \frac{j+2}{n+2}\frac{j+1}{n+1}(
s_{AB/Ab}i + s_{Ab/aB}k + s_{Ab/ab}(n-i-j-k) - s_{Ab/Ab}(n-j)
)\Psi_{n+2}(i, j+2, k) \\\nonumber
& + \frac{j+1}{n+2}\frac{k+1}{n+1}(
  s_{AB/Ab}i
  + s_{AB/aB}i
  + s_{Ab/Ab}j
  + s_{Ab/ab}(n-i-j-k) \\\nonumber & \hspace{20pt}
  + s_{aB/aB}k
  + s_{aB/ab}(n-i-j-k)
  - s_{Ab/aB}(2n - j - k)
)\Psi_{n+2}(i, j+1, k+1) \\\nonumber
& + \frac{k+1}{n+2}\frac{n-i-j-k+1}{n+1}(
  s_{AB/aB}i
  + s_{AB/ab}i
  + s_{Ab/aB}j
  + s_{Ab/ab}j \\\nonumber & \hspace{20pt}
  + s_{aB/aB}k
  - s_{aB/ab}(n + i + j)
)\Psi_{n+2}(i, j, k+1) \\\nonumber
& + \frac{k+2}{n+2}\frac{k+1}{n+1}(
  s_{AB/aB}i
  + s_{Ab/aB}j
  + s_{aB/ab}(n-i-j-k)
  - s_{aB/aB}(n - k)
)\Psi_{n+2}(i, j, k+2).
\end{align}
Multiplying through by $2N_{ref}$ gives us selection operators in terms of
$\gamma$ instead of $s$.

# Data analysis

## DFE for missense and LOF variants

@@

## Multinucleotide mutations and positive LD between linked synonymous variants

@@

In fitting the LD decay of $\sigma_d^1$, the best fit parameters were
$A=0.132$, and $\lambda=0.0103$. An exponential scaling of $0.01$ implies that
the vast majority of new mutations *pairs* do not occur via MNMs for distances
greater than 200 bp, though a substantial fraction ($10-15\%$) occur via MNMs
for very tightly linked loci with distances on the order $0-50$ base pairs. It
is important to note that this does not mean that $10-15\%$ of new mutations
occur via MNMs, since this fraction is contitioned on two mutations occurring
at short distances.

## Grouping Thousand Genomes populations based on clustering

The large confidence intervals for measurements of signed LD could be driven by
either averaging over relatively few observed pairs of mutations, or due to
small sample sizes that make each individual measurement noisy estimates of the
LD for that pair of mutations in the full population. To explore the underlying
cause of measurement uncertainty in the @1000_Genomes_Project_Consortium2015-zq
data, I considered larger sets of samples by combining populations that
consistently cluster together in PCA and UMAP space and have low
differentiation [@Diaz-Papkovich2020-ee]. I took combinations of CEU/GBR,
CHB/CHS, CDX/KHV, and MSL/GWD. While recognizing that residual population
structure in these population combinations could alter expected LD statistics
compared to the respective single-population estimates, I was more interested
in the effect that increasing the sample sizes would have on estimated
measurement error.

Across each of the four combinations tested, confidence intervals were roughly
equivalent to those of each of the individual populations. This suggests that
the limiting factor to accurate LD measurement is not sample size but rather
the overall levels of diversity and number of pairs of mutations that we
compare. $\mathbb{E}[D]$ is most affected by common variants, and the sample
sizes of the Thousand Genomes Project data are likely sufficient to accurately
estimate common allele frequencies. Adding additional samples will increase the
number of rare variants that we observe, but rare variants have minimal impact
on $\sigma_d^1$. Thus, the accuracy of estimates of $\sigma_d^1$ is more
fundamentally limited by evolutionary history and genome biology (i.e. past
population sizes, mutation and recombination rates) than by sample sizes.
